---
title: "rROMA"
author: "Luca Albergante (luca.albergante@gmail.com)"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
---

This package provides an R implementation of [ROMA](http://journal.frontiersin.org/article/10.3389/fgene.2016.00018/full). The package is under active development and is currently not stable.

A Java implementation developed by Andrei Zynovyev and is also [available](https://github.com/sysbio-curie/Roma).

## Intalling rROMA

The rRoma package relies on the `scater` package, which is available only on BioConductor. This package can be installed with the following command

```{r, eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("scater")
```

rRoma ccan then be installed using `devtools`

```{r, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("Albluca/rROMA")
```

## Using rROMA

The package can be loaded with the usual syntax, i.e., by typing

```{r}
library(rRoma)
```

rRoma requires a gene expression matrix - with column names indicating samples and row names indicating gene names - and a module file containing information on the genesets that need to be evaluated. The module file can be loaded from a GMT file. Functions to automate the generation of then GMT file are also available.

```{r}



```

